Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-04-05 Manuscript/manuscript.Rmd
manuscript edits

We first examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased. That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins (Supplementary Figure 15 (A)). The imbalance ratios, however, remained relatively stable (Supplementary Figure 15 (B)). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and minority classes (i.e. a genomic bin was much more likely to not overlap with a flanked TAD boundary than it was to overlap with one). Therefore suitable resampling techniques were needed.

@@

After suitably reducing the feature space, we performed individual random forest classification models specific to cell line, data resolution, resampling technique, and predictor type. We assessed each models' ability to correctly identify the genomic bins that overlap with flanked TAD boundaries using F1-Score, MCC, and balanced accuracy. For 10 kb data resoltion, it was found that models with distance-type predictors performed better according to all metrics, across each cell line, regardless of resampling technique used (Figure 4). This result was consistent across each of the other data resolutions as well (Supplementary Figure 4). Notably, we saw marked improvements in performance for models implemented on resampled datasets when compared to no resampling.

@@

We then directly compared models using SMOTE resampled datasets across each data resolution to see how performances changed as resolution increased. For each cell line, we found that model performances tended to decline as data resolution increased. Random forest models implemented on 10 kb, SMOTE resampled, data resolutions, with distance-type predictors were found to perform best according to MCC and balanced accuracy (Figure 5). There was a slight improvement in F1-Score for the 25 kb resolution on the GM12878 cell line. Performances from the K562 cell line degraded much faster as resolution increased, while the worst performing models for the GM12878 cell line were at the 50 kb resolution, followed by a slight improvement in performance at the 100 kb resolution. To be comprehensive we evaluated the performances across resolutions of models with distance-type predictors for the other resampling techniques and found the results to be largely consistent (Supplementary Figure 5).

## DATA RESAMPLING IMPROVES MODEL PREDICTION

>>>>

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased. That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins (Supplementary Figure 15 (A)). The imbalance ratios, however, remained relatively stable (Supplementary Figure 15 (B)). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and minority classes (i.e. a genomic bin was much more likely to not overlap with a flanked TAD boundary than it was to overlap with one). Therefore suitable resampling techniques were needed.

@@

We first evaluated model performances using only the overlap counts predictor types without re-sampling as these characteristics represented models with the least granularity and could act as a baseline. We measured the performances across resolutions for each cell line (Figure 7 (A)). Some notable results became clear. First, we see that the conventional AUC value was unable to accurately portray the poor performance of the models when compared to the composite metrics F1-score, MCC, and, to a lesser extent, the balanced accuracy. Inspecting the F1-score and MCC more closely, we see that some models at lower resolutions (for either cell line) were unable to correctly predict any bins that overlap with flanked TAD boundaries in the test data, as indicated by the missing bars.

## DATA RE-SAMPLING IMPROVES MODEL PERFORMANCE

Comparisons of the effects of the various resampling techniques were then made across resolutions, for each cell line. Results showed that performances drastically improved among the composite metrics (Figure 7 (B-D)). Models implemented on randomly under-sampled and SMOTE-sampled data performed similarly, while performances on randomly over-sampled data lagged behind, yet still outperformed models with no data re-sampling. Overall, there appeared the be a downward trend in model performance as resolutions decreased, across both cell lines.

## USING DISTANCE-TYPE PREDICTORS FURTHER IMPROVED MODEL PERFORMANCES

After illustrating the improvement in model performance using re-sampled data for only the overlap counts predictor types, we then directly compared the performances of each re-sampling techniques on the other two predictor types (overlap percents and distances). Immediately, results showed that models with distance-type predictors outperformed models with other predictor types (Figure 8 (A-D)). This results was consistent across resolution and cell line. As before, we are able to gain much more insight into the performances of the models by using composite metrics as opposed to simply the AUC, in cases of class imbalance. Likewise, performances were shown to continue to degrade as resolutions decreased.

## MODELS IMPLEMENTED ON SMOTE RE-SAMPLED DATA WITH DISTANCE-TYPE PREDICTORS PERFORMED BEST AMOUNG ALL ENSEMBLE SYSTEMS

Of the re-sampling techniques considered, it was found that SMOTE re-sampled datasets, implemented on distance-type predictors, yielded the highest model performances across each of the 3 composite metrics (Figure 9). These results were consistent across resolution and cell line. Although RUS performed only marginally worse than SMOTE, there was a steap drop off in performances for ROS and no re-sampling, across both resolution and cell line. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

@@

Furthermore, we found that model performances tended to decline as data resolution decreased. Random forest models implemented on 10 kb, SMOTE resampled, data resolutions, with distance-type predictors were found to perform best according to MCC and balanced accuracy. There was a slight improvement in F1-Score for the 25 kb resolution on the GM12878 cell line. Performances from the K562 cell line degraded much faster as resolution decreased, while the worst performing models for the GM12878 cell line were at the 50 kb resolution, followed by a slight improvement in performance at the 100 kb resolution.
